Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-04 paper/main.tex
start working on section 2

Available techniques split in two categories.  Tools
from the first one derive impressive bounds for
imperative programs, but they are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no opportunity for sound interaction with the
tool, for example, manual proofs cannot be composed
with automatically inferred bounds.  One reason for this
impossibility to interact is that available tools work
on a highly abstract version of the source program. This
abstracted program has a lot more behaviors than the
original one and is usually expressed in a language that
is not able to express memory operations.

@@

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  It handles, using
only one simple mechanism:

@@

>>>>

Available techniques split in two categories.  Tools
from the first one derive impressive bounds for
imperative programs, but they are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no opportunity for sound interaction with the
tool, for example, manual proofs cannot be composed
with automatically inferred bounds.  One reason for
this impossibility to interact is that available tools
work on an abstract version of the source program. This
abstracted program has a lot more behaviors than the
original one and is usually expressed in a language that
does not feature memory operations.

@@

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

@@

\sectskip
\section{Overview and Examples}
\aftersectskip

In this section, we informally introduce the automatic amortized
analysis (AAA) for Clight programs.  Some example bounds are derived
step by step following our algorithm.

\subsection{Bounding resource usage with potential functions}

The idea that underlies the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes
on a starting state $\state$ and consumes $n$ resource units of
some user-defined quantity.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after
the execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a program
state such that $(S, \state) \Downarrow_n \state'$.  This gives that
$\Phi(\state)$ is a valid resource bound.

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}.
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

The considerations above already show a departure from classical
techniques for imperative programs.  Indeed, reasoning with two
potential functions is more compositional than using
ranking-functions because it forces to consider the sequencing of program
operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though
it was derived on $S_1$ only.

% Potential functions can access the whole program state, so this
% framework can readily be used to derive bounds that depend on the
% memory contents.

The reasoning presented above is very general and flexible.  But in
practice, to make the reasoning tractable on large code bases, we would
like to infer automatically the potential functions.  The first step
towards this goal is to reduce our search space by fixing the shape
of potential functions $\Phi$.  We settle on a form that allows
\emph{linear programming} (LP) to help inference:
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
  q_{(x,y)} \cdot |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  We assumed
that the program state is a simple map from program variables to integers
and we treat constants as program variables (that cannot be changed
by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
the sum.  We then develop an inference rule for each syntactic construct
that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
the previous section.

This idea is best explained by example.  If we use the tick metric that
assigns cost $n$ to the function call \code{tick(n)} and cost $0$ to all
other operations then the cost of the following example can be bounded
by $|[x,y]|$.\footnote{Note that there are no restrictions on the signs
of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start with
the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
= |[\state(x),\state(y)]|$) which we also use as the loop invariant.
For the loop body we have (like in Hoare logic) to derive a triple
like $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do
so if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential $|[x,y]|$
and the fact that $|[x,y]| > 0$ before the assignment.  If we denote the
updated version of $x$ after the assignment by $x'$ then the relation
$|[x,y]| = |[x',y]| + 1$ between the potential before and after the
assignment \code{x=x+1} holds.  This means that we have the potential
$|[x,y]| + 1$ before the statement \code{tick(1)}.  Since \code{tick(1)}
consumes one resource unit, we end up with potential $|[x,y]|$ after
the loop body and have established the loop invariant again.

\subsection{Compositionality in practice}

The following example is typical in cryptographic primitives
implementations.  These primitives work by block, so when data of
arbitrary length needs to be processed, it is munched in blocks and
the leftovers are stored in a buffer for future use when more data
is available.  This looping pattern is particularly well handled by
our tool because the LP solver is leveraged to generate a tight bound.
The optimization problem it is given encodes that constant bounds are
preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
AAA is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
counted as constant.
%
\begin{lstlisting}
for (; l >= 8; l -= 8)
    // process one block
    tick(N);
for (; l > 0; l--)
    // save leftovers
    tick(1);
\end{lstlisting}
%
It is worth taking some time to explain the constant contribution
to the resource bound when $N<8$.  Remark first that the cost of
the second loop is $|[0,l]|$, and after the first one, we still have
$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
potential units instead.  This explains the somewhat intriguing value
of the constant term in the bound.  All this reasoning is done within
our tool automatically and simply relies on general rules presented
in further sections.

This level of precision and compositionality is only achieved by
our work, no other available tool will attain the above tight bounds
on this common kind of looping code.  Thanks to the modularity of
amortized analysis the intricate interactions between the bounds of
the two loops can be modeled as one linear program, and because LP
solvers can optimize the quality of the final bound we can keep
it close to the actual resource consumption.
